---
title: "Rotation Data Exploration"
output: html_document
---

Compiling/exploring clinical data avaiable from subjects with 7T data to narrow down rotation project ideas.
NOTE: Code pulls data saved in directory `/data` not tracked in git

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load packages

```{r Install/Load packages, echo=FALSE, message=FALSE, warning=FALSE}
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(tidyverse, ggplot2)
```

## Clinical Data

### PNC1 Data 

Matching to see how many 7T glucest participants are part of PNC

Load glucest subject list extracted from .xls (list from David 1/13/22)

```{r load glucest}
glucest <- read.csv("data/7t_dates.csv", header=T) 
names(glucest)
str(glucest)
count(glucest)
```

Loading PNC subject list (PMACS LPC mapped)

``` {r load pnc dx}
pnc <- read.csv("/Users/megardn/ImageData/PMACS_remote/clinical/n1601_diagnosis_dxpmr_20170509.csv", header=T)
names(pnc)
str(pnc)
```

join dataframes
```{r join, echo=FALSE}
merge <- inner_join(glucest, pnc,
  by = c("BBLID"="bblid"),
  copy = FALSE,
  keep = FALSE,
  na_matches= "never")
n_distinct(merge$BBLID) #count unique ppts in both
names(merge)
```

### RedCap Data 

Combining data from Kosha from later RedCap studies with PNC1 data from LPC

Diagnoses across time for 7T participants

``` {r load.redcap}
#load in longitudinal redcap diagnosis data
dx <- read.csv("data/diagnosis_long.csv", header = TRUE, stringsAsFactors = F, na = "")
names(dx)
```

``` {r}
redcap <- dx %>% select(BBLID, DODIAGNOSIS, HSTATUS, AXIS1_DX1, allaxis1desc, GASR_CURRENT, AGEONSET_AXIS1, AGEONSET_CLINICRISK, dx_pscat, timepoint)
names(redcap)
#drop 22q subjects
redcap_no22 <- subset(redcap, HSTATUS != "22q" | is.na(HSTATUS))
```

### Diagnoses

Initialize a dataframe to store counts as we go
``` {r}
df_sum <- data.frame(Variable=character(),
                     Frequency=integer(),
                     Note=character(),
                     stringsAsFactors=F)
```

find subjects with multiple CAPA diagnoses

``` {r}
length(which(redcap_no22$timepoint==2))

#append to summary df
df_sum[nrow(df_sum) + 1,] = c("CAPA", length(which(redcap_no22$timepoint==2)), "subjects with 2+ diagnoses in REDCap (CAPA)")
```

76 subjects have 2 or more dx only including CAPA data

combine dataframes
```{r}
#drop the PNC data down to columns that will make sense going forward (e.g. only clinical info at that timepoint)
merge.simple <- merge %>% subset(select = -c(scanid, X7T_date, ONM_7T_date))
#make each df have the same number/names of columns
empty_cols <- c("goassessDxpmr4", "goassessDxpmSep15", "goassessDxpmr6", "goassessDxpmr7") 
redcap2 <- redcap_no22 #copy dataframe before adding empty columns
redcap2[ , empty_cols] <- NA
empty_cols2 <- c("DODIAGNOSIS", "HSTATUS", "AXIS1_DX1", "allaxis1desc", "GASR_CURRENT", "AGEONSET_AXIS1", "AGEONSET_CLINICRISK", "dx_pscat")
merge.simple[ , empty_cols2]<- NA
merge.simple[ , "timepoint"] <- 0 #no dates for PNC dx, setting timepoint to 0

names(redcap2)
names(merge.simple)

total <- rbind(redcap2, merge.simple)
total
```


find numbers - shouldn't be any overlaps between dx listed in redcap and dx from PNC1 since they're different instruments

``` {r counts}
uniq <- data.frame(table(total$BBLID)) #find frequency of each ID in dx list
names(uniq)
uniq <- uniq %>% rename(BBLID=Var1) #rename variable

sum(is.na(uniq$BBLID))
nrow(uniq) #number of unique subjects from 7T dataset with diagnoses in PNC and/or redcap

#append to summary df
df_sum[nrow(df_sum) + 1,] = c("unique_dx", nrow(uniq), "number unique subjects with diagnoses in PNC Goassess and/or REDCap CAPA")
```

108 unique subjects with diagnoses (PNC and/or redcap)

```{r followups}
follow <- filter(uniq, Freq > 1)
summary(follow)
nrow(follow)

#append to summary df
df_sum[nrow(df_sum) + 1,] = c("multi_dx", nrow(follow), "number unique subjects with 2+ diagnoses in PNC Goassess and/or REDCap CAPA")
```

84 subjects with 7T scans have 2 or more diagnoses between PNC or RedCap

Find initial diagnoses of subj with 2+ scans
``` {r}
#sort diagnoses by order of interview and filter down to earliest dx only
first <- total[order(total$timepoint),] %>% 
  group_by(BBLID) %>% 
  distinct(BBLID, .keep_all=TRUE) %>% 
  select(BBLID, HSTATUS, allaxis1desc, dx_pscat, goassessDxpmr4) %>% 
  mutate(HSTATUS = as.factor(HSTATUS), dx_pscat = as.factor(dx_pscat), goassessDxpmr4 = as.factor(goassessDxpmr4))

summary(first)

#extract dx frequencies
pscat_freq <- as.data.frame(table(first$dx_pscat))
goassess_freq <- as.data.frame(table(first$goassessDxpmr4))
dx_freqtotal <- rbind(pscat_freq, goassess_freq) %>% rename(dx=Var1)

#combining overlapping dx categories across goassess/recap & aggregate across
#NOTE!!! 4PS goassess group may also include some prodrome
levels(dx_freqtotal$dx)
dx_freqtotal$dx <- fct_collapse(dx_freqtotal$dx, "healthy" = c("noDSMdx", "1TD"),
                                "other" = c("2WK", "other"),
                                "CHR" = c("pro","3PC"),
                                "psych" = c("psy", "4PS"))

#aggregate across
dx_freqtotal <- dx_freqtotal %>% group_by(dx) %>% summarize(Freq=sum(Freq))
dx_freqtotal
```

Now looking for participants who 2+ diagnoses, at least one of which occurred after their matched 7T scan - must be CAPA, all redcap should be pre-7T
```{r}
dx_multi <- total %>% subset(BBLID %in% follow$BBLID) #filter dx data to only subj with 2+ timepoints
length(unique(dx_multi$BBLID)) #confirm filtered correctly

#reorganize diagnoses to best compare across goassess and CAPA
dx_multi <- dx_multi %>% unite(DIAGNOSIS, c (dx_pscat, goassessDxpmr6), na.rm = TRUE, remove=FALSE) %>% 
  mutate(DIAGNOSIS = as.factor(DIAGNOSIS),
         DIAGNOSIS = fct_collapse(DIAGNOSIS, "TD" = c("noDSMdx", "TD"),
                                "other" = c("OP", "other"),
                                "GO.PS" = c("PS"),
                                "CAP.pro" = c("pro", "psy")))
levels(dx_multi$DIAGNOSIS)

#get 7T dates Kosha used and set formats
date7t <- read.csv("data/margaret_7T_series.csv", header = TRUE, stringsAsFactors = F, na = "") %>% mutate(date_7T = as.Date(date_7T, format ="%m/%d/%y"))

#merge and set date format, arbitrarily set PNC dx dates to 1/1/2000
dx_multi <- inner_join(dx_multi, date7t, by = c("BBLID" = "bblid")) %>% 
  mutate(DODIAGNOSIS= ifelse(is.na(DODIAGNOSIS), "2000-01-01", DODIAGNOSIS),
         DODIAGNOSIS = as.Date(DODIAGNOSIS))
str(dx_multi$DODIAGNOSIS)

#calc diff (negative values -> dx before 7t)
dx_multi$diff_in_days = as.numeric(difftime(dx_multi$DODIAGNOSIS, dx_multi$date_7T, units = "days"))

#find subjects that have a dx >356 days post-7T
post7t <- dx_multi %>% filter(diff_in_days >= 365)
n_distinct(post7t$BBLID)

#append to summary df
df_sum[nrow(df_sum) + 1,] = c("1yr_dx", n_distinct(post7t$BBLID), "number unique subjects with 2+ diagnoses in PNC Goassess and/or CAPA who have a dx >= 365 days post-7T")
```

Some people have multiple 7T scan dates - redo by using earliest 7T date to see if there are more ppl with a 1 yr + clinical follow up
``` {r}
#start with complete glucest date list - clean up and compare X7T_date and ONM_7T_date
str(glucest)
#replace blank values with na, get rid of extra characters in X7T_date
first7t <- glucest %>% mutate_all(na_if,"") %>%
  mutate(X7T_date = gsub(",.*","", X7T_date), 
         X7T_date = as.Date(X7T_date,format ="%m/%d/%y"),
         ONM_7T_date = as.Date(ONM_7T_date,format ="%m/%d/%y"))
str(first7t)

#create new column with only first scan dates
first7t <- first7t %>% mutate(first_7t_date = pmin(X7T_date, ONM_7T_date, na.rm=TRUE)) %>% select(BBLID, first_7t_date)

#merge into clinical data and drop scan dates from earlier
dx_multi_first7t <- inner_join(dx_multi, first7t, by = c("BBLID" = "BBLID")) %>% 
  subset(select = -c(date_7T, diff_in_days)) %>%
  mutate(BBLID = as.factor(BBLID))
#calc diff (negative values -> dx before 7t)
dx_multi_first7t$diff_in_days = as.numeric(difftime(dx_multi$DODIAGNOSIS, dx_multi_first7t$first_7t_date, units = "days"))

#find subjects that have a baseline dx up to 364 days post-7T
clin364 <- subset(dx_multi_first7t, diff_in_days <= 364, select = BBLID) %>% unique()
#find subjects that have a followup dx >365 days post-7T
clin365 <- subset(dx_multi_first7t, diff_in_days >= 365, select = BBLID) %>% unique()
#filter df
post_first7t <- subset(dx_multi_first7t, (BBLID %in% clin365$BBLID & BBLID %in% clin364$BBLID))

length(unique(post_first7t$BBLID))
levels(post_first7t$DIAGNOSIS)

#append to summary df
df_sum[nrow(df_sum) + 1,] = c("1yr_dx_earliest7t", nrow(clin365), "number unique subjects with 2+ diagnoses in PNC Goassess and/or CAPA who have a baseline dx (up to 364 d) and follow-up dx (>= 365 d) after their earliest 7T")
```

32 subjects have clinical data after their earliest 7T!

plot a timeline of each subject's clinical diagnoses in relation to scan date

```{r}
ggplot(post_first7t, aes(x=diff_in_days, y=BBLID, color=DIAGNOSIS)) + 
  geom_point() +
  geom_vline(xintercept=365) +
  labs(title="Diagnosis Relative to First 7T Scan", x="Days from 7T Scan")
```
see diagnostic trajectory across time -> binned
``` {r fig.height=7}
#make sure timepoints are factors in an order that makes sense
levels(post_first7t$timepoint)
post_first7t$timepoint<- ordered(post_first7t$timepoint, levels=c("0", "1", "2", "3", "4", "5", "6", "7", "8"))

dx_7t_wide <- post_first7t %>% select(BBLID,DIAGNOSIS, timepoint, first_7t_date) %>%
  arrange(timepoint) %>% 
  pivot_wider(names_from=timepoint, names_prefix = "timepoint_", values_from=DIAGNOSIS) %>%
  unite(trajectory, c(3:10), na.rm = TRUE, remove=FALSE) %>% 
  mutate(trajectory = as.factor(trajectory))

#plot
ggplot(dx_7t_wide) + 
  geom_bar(aes(x=trajectory), stat="count") + 
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("frequency of categorical trajectories")
```

find change in dx across time - this includes changes before 7t and after 7t - want to see just changes where the 7t happened in the middle

``` {r}

#finding ppt where dx changed between timepoints
dx_changes <- post_first7t %>%
  group_by(BBLID) %>%
  arrange(BBLID, diff_in_days) %>%
  mutate(dx_change = case_when(DIAGNOSIS != lag(DIAGNOSIS) ~ TRUE, TRUE ~ FALSE))

dxchange_ppt <- subset(dx_changes, dx_change == "TRUE", select = BBLID) %>% unique()
nrow(dxchange_ppt)

#ppts who's first dx is different than last dx (know they have at least one assess on either side of 7t date, so should capture change over time)
first_last <- post_first7t %>% 
  group_by(BBLID) %>% 
  arrange(DODIAGNOSIS) %>% 
  slice(c(1,n())) %>%
  mutate(dx_change = case_when(DIAGNOSIS != lag(DIAGNOSIS) ~ TRUE, TRUE ~ FALSE)) %>%
  ungroup()

subset(first_last, dx_change == "TRUE", select = BBLID) %>% nrow()

#same count - is it a problem if change dx occured prior to scan???
```

get frequency of each initial dx

```{r}
first_only <- first_last %>% group_by(BBLID) %>% 
  arrange(DODIAGNOSIS) %>% 
  slice(c(1))

last_only <- first_last %>% group_by(BBLID) %>% 
  arrange(DODIAGNOSIS) %>% 
  slice(c(-1))

table(first_only$DIAGNOSIS) 
table(last_only$DIAGNOSIS) 
```
### GAF

first looking at GASR data included in diagnosis_long.csv

``` {r gasr}
#drop missing gasr
gasr <- redcap_no22 %>% drop_na(GASR_CURRENT)

#list subj with >=2 gasr scores
multi_gasr <- gasr %>% subset(timepoint==2) %>% select(BBLID)
#merge to filter to only subj with multiple gasr
gasr_long <- inner_join(gasr, multi_gasr, by = c("BBLID"))

#filter to first and last gasr scores for each subject
gasr_long <- gasr_long[order(gasr_long$timepoint),] %>% group_by(BBLID) %>% slice(c(1,n()))

#calculate change in GASR, save and plot
gasr_change <- gasr_long %>% group_by(BBLID) %>% mutate("gaf_diff"= GASR_CURRENT - lag(GASR_CURRENT), BBLID=as.factor(BBLID)) %>% select(BBLID, gaf_diff) %>% na.omit 
summary(gasr_change$gaf_diff)

#append calculation of gasr to summary df
df_sum[nrow(df_sum) + 1,] = c("multi_gasr", nrow(gasr_change), "number unique subjects with 2+ GASR values from REDCap diagnosis.csv")

#plot change in gasr
ggplot(gasr_change, aes(x=gaf_diff)) + 
  geom_histogram(bins=25, color="white", fill= "darkgreen") + 
  labs(title="change in GASR")
```

looking through SIPS GAF current

``` {r gafc}
#load SIPS
sips <- read.csv("data/sips_long.csv", header = TRUE, stringsAsFactors = F, na = "")
summary(sips)

#make list of 22q subj and drop
q22 <- subset(redcap, HSTATUS == "22q", select = BBLID) %>% unique()
sips_no22 <- subset(sips, !(bblid %in% q22$BBLID))

#drop missing gasfc
gafc <- sips_no22 %>% drop_na(GAF_C) 

#list subj with >=2 gafc scores
multi_gafc <- gafc %>% subset(timepoint==2) %>% select(bblid)
#merge to filter to only subj with multiple gafc
gafc_long <- inner_join(gafc, multi_gafc, by = c("bblid"))

#filter to first and last gasr scores for each subject
gafc_long <- gafc_long[order(gafc_long$timepoint),] %>% group_by(bblid) %>% slice(c(1,n()))

#calculate change in GASR, save and plot
gafc_change <- gafc_long %>% group_by(bblid) %>% mutate("gafc_diff"= GAF_C - lag(GAF_C), BBLID=as.factor(bblid)) %>% select(bblid, gafc_diff) %>% na.omit 
summary(gafc_change$gafc_diff)

#append calculation of gasr to summary df
df_sum[nrow(df_sum) + 1,] = c("multi_gafc", nrow(gafc_change), "number unique subjects with 2+ GAF Current values from SIPS")

#plot change in gasr
ggplot(gafc_change, aes(x=gafc_diff)) + 
  geom_histogram(bins=25, color="white", fill= "darkgreen") + 
  labs(title="change in GAF Current")
```


## Outputs
save summary document
``` {r summary.output}
write.csv(df_sum, file = "7T_ppt_clinical_summary.csv")
print(df_sum)
```

## To Do:
clean up dx_multi, get rid of attempt with kosha's 7T dates
see SIPS timepoints with 7T
look at PRIME

plot all dx across time in categories
plot change in dx across most proximal scans

