---
title: "GluCEST-SOPS-Prediction"
output: html_document
---

```{r Install/Load packages, echo=FALSE, message=FALSE, warning=FALSE}
#load packages
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(tidyverse, ggplot2, ggpubr, ggrepel, skimr, lubridate, tidyr, data.table, car, ggpubr)
```

**NOTES TO SELF**
We want the final df to be long, not wide


# Compile Scan Lists
*from `scan_list_comp.Rmd`*
Compile lists of all GluCEST scans.

**To DO:**
- confirm that newer scan dates = Terra 
- pull scan IDs

## Load & Clean Lists

7T list from David (up to ~ summer 2019)

```{r 7t list}
glu_old <- read.csv("data/7t_dates.csv", na.strings = "") # read in 7T list from David (up to ~ summer 2019)
summary(glu_old)

# clean up formatting
glu_old <- glu_old %>%
  mutate(BBLID=as.character(BBLID),
  X7T_date = gsub(",.*","", X7T_date), 
  X7T_date = as.Date(X7T_date,format ="%m/%d/%y"),
  ONM_7T_date = as.Date(ONM_7T_date,format ="%m/%d/%y")) %>%
  rename(Terra=X7T_date,
         ONM=ONM_7T_date)
# make long
glu_old.long <- gather(glu_old, scanner, "7T_date", Terra:ONM, factor_key=TRUE) %>%
  drop_na("7T_date")
```

7T list from Arianna's LongGluCEST study (pulled April 2022)

``` {r longglucest}
glu_ar <- read.csv("data/longglucest_scans.csv", header=T) # read in 7T list from Arianna LongGluCEST study (pulled April 2022)
summary(glu_ar)

#clean up formatting
glu_ar <- glu_ar %>%
  mutate(BBLID=as.character(BBLID),
         baseline_clinical = as.Date(baseline_clinical, format ="%m/%d/%y"),
         base_7t = as.Date(base_7t,format ="%m/%d/%y"),
         follow_7t = as.Date(follow_7t,format ="%m/%d/%y"))

# make long
glu_ar.long <- gather(glu_ar, scanner, "7T_date", base_7t:follow_7t, factor_key=TRUE) %>%
  drop_na("7T_date") %>%
  mutate(scanner= fct_collapse(scanner, Terra = c("base_7t", "follow_7t")))
```

7T list from Heather's Aging study

``` {r aging}
glu_age <- read.csv("data/7tglucestage_scans.csv", header=T) # read in 7T list from Heather Aging study (pulled April 2022)

#clean up
glu_age <- glu_age %>%
  transmute(BBLID=as.character(BBLID),
         "7T_date" = as.Date(DOSCAN,format ="%m/%d/%y"),
         scanner = as.factor("Terra"),
         SCANID=as.character(SCANID))
```

Join all lists:

```{r}
#join long versions of old and Longitudinal scan lists
long1 <- merge(x=glu_old.long, y=glu_ar.long,by=c("BBLID", "7T_date", "scanner"),all=TRUE)
str(long1)

#join in Age study scan list
all_7t.long <- merge(x=long1, y=glu_age, by=c("BBLID", "7T_date", "scanner"), all=TRUE)
all_7t.long <- all_7t.long %>% rename(date_7t = "7T_date")
str(all_7t.long)

dup <- duplicated(all_7t.long[,1:2])
any(dup) #no duplicates!
```

Next make a wide version to easily add in clinical/subject-level data.

``` {r}
all_7t.count <- all_7t.long %>%
  group_by(BBLID) %>%
  arrange(all_7t.long$date_7t) %>%
  mutate(count = as.character(row_number(BBLID))) %>%
  ungroup() # add counts
all_7t.wide <- pivot_wider(all_7t.count, id_cols=BBLID, names_from = count, values_from=c(date_7t, scanner, SCANID)) # pivot
colnames(all_7t.wide)[2:ncol(all_7t.wide)] <- paste("scan", colnames(all_7t.wide[,c(2:ncol(all_7t.wide))]), sep = "_") #rename cols

# #write final list to csv
# write.csv(all_7t.wide, "data/all_7T_april22.csv", row.names=FALSE, na="")
# save(all_7t.wide, file="data/all_7T_april22.Rdata")

n_distinct(all_7t.wide$BBLID)
```

Load diagnosis data from oracle ID and remove 22q subj

```{r}
#load diagnosis data
redcap <- read.csv("data/oracle_diagnoses_all.csv", header = TRUE, stringsAsFactors = F, na = c("", "null")) %>% 
  select(BBLID, DODIAGNOSIS, HSTATUS, AGEONSET_CLINICRISK | contains("AXIS1_")) %>% #get manageable subset of data
  mutate(DODIAGNOSIS = as.Date(DODIAGNOSIS),
         BBLID=as.factor(BBLID))


#make list of 22q subj
q22 <- subset(redcap, HSTATUS == "22q", select = BBLID) %>% unique()

#remove 22q participants from scan list, that way it'll apply to all subsequent merges w/ clinical data
all_7t.wide <- all_7t.wide %>% subset(!(BBLID %in% q22$BBLID)) %>% as.data.frame()#drop 22q sub
all_7t.long <- all_7t.long %>% subset(!(BBLID %in% q22$BBLID)) %>% as.data.frame()#drop 22q sub
```

There are `r length(unique(all_7t.wide$BBLID))` (197) unique subjects who have 7T scans, excluding 22q pts ID'd from Kosha's RedCap data.

# SIPS/SOPS

Load SIPS data

```{r}
#load SIPS
sips <- fread("data/oracle_sips_all.csv", na.strings="null", header=TRUE)
  #mutate(DOSIPS = as.Date(DOSIPS),
         #bblid = as.factor(bblid))
sips$BBLID <- as.factor(sips$BBLID)
skim(sips)
```

Now for our outcome of interest, SIPS/SOPS scores. First filtering by completion. 

```{r}
#list sops items
plist <- c("P1", "P2", "P3", "P4", "P5")
nlist <- c("N1", "N2", "N3", "N4", "N5", "N6")
dlist <- c("D1", "D2", "D3", "D4")
glist <- c("G1", "G2", "G3", "G4")
sops.item.list <- c(plist, nlist, dlist, glist)

#drop missing sops items
sips_filt <- sips %>% drop_na(c(sops.item.list))
#sum(is.na(sips_filt[,c(sops.item.list)]))
length(unique(sips_filt$BBLID))
```

Some subjects have multiple assessments on one day (one from proband one from family, etc); need to be sure to pick only `TYPE == "COMBINED"` per discussion with David. If no combined, pick proband.

```{r}
#if subj has more than one assessment for same day, select Combined 
sips_filt<- sips_filt %>% mutate(TYPE = as.factor(TYPE))
levels(sips_filt$TYPE) 
sips_filt$TYPE <- fct_collapse(sips_filt$TYPE, "Proband" = c("Proband", "proband", "P", "FP", "IP"), "Combined"=c("Combined", "FPC")) #combine all proband
sips_filt$TYPE <- factor(sips_filt$TYPE, levels = c("Combined", "Proband", "Collateral"))

#group data to select correct type (defined as lowest ordered level) - also cleaning up
sips_filt2 <- sips_filt %>%
  group_by(BBLID, DOSIPS) %>%
  arrange(TYPE) %>%
  slice(c(1)) %>%
  ungroup() %>%
  select(-SIPS_SOURCEID, -SOURCE_PROJECT) %>%
  mutate(DOSIPS=as.Date(DOSIPS))

length(unique(sips_filt2$BBLID)) #confirm no subjects deleted
skim(sips_filt2)
```

Calculate SOPS scores by summing within each subscale (seems to be valid metric based on clinicaltrials.gov)

```{r, warning=FALSE}
#calculate sops scores
sips_scored<- sips_filt2 %>% mutate(sops_p = rowSums(across(plist), na.rm=F),
                                  sops_n = rowSums(across(nlist), na.rm=F),
                                  sops_d = rowSums(across(dlist), na.rm=F),
                                  sops_g = rowSums(across(glist), na.rm=F),
                                  sops_tot = rowSums(across(c(sops_p, sops_n, sops_d, sops_g), na.rm=F)))

#also dropping `3456` cols (and maybe GAF_PCTCHG?) since they don't seem to mean anything or be useful and it would be better to recalculate them anyway
#aksi dropping rater
sips_scored <- sips_scored %>% select(-VISITNUM, -RATER, -REVIEW, -POS_345, -POS_6, - NEG_3456, -DIS_3456, - GEN_3456) %>% as.data.frame()

head(sips_scored)
```


Now merge SOPS in with scans

```{r}
sips.and.scans <- inner_join(all_7t.wide, sips_scored, by = "BBLID")
length(unique(sips.and.scans$BBLID))

#calc diff from each 7T scan (negative values -> dx before 7t)
sips.and.scans <- sips.and.scans %>%
  mutate(sips_diff_days1 = as.numeric(difftime(DOSIPS,scan_date_7t_1, units = "days")),
         sips_diff_days2 = as.numeric(difftime(DOSIPS,scan_date_7t_2, units = "days")),
         sips_diff_days3 = as.numeric(difftime(DOSIPS,scan_date_7t_3, units = "days")))

length(unique(sips.and.scans$BBLID)) #no subj lost

#label SIPS that are baseline or post for each scan
sip.pre.scan1 <-  which(sips.and.scans$sips_diff_days1 <= 364 & sips.and.scans$sips_diff_days1 >= -365)
sip.pre.scan2 <-  which(sips.and.scans$sips_diff_days2 <= 364 & sips.and.scans$sips_diff_days2 >= -365)
sip.pre.scan3 <-  which(sips.and.scans$sips_diff_days3 <= 364 & sips.and.scans$sips_diff_days3 >= -365)
sip.post.scan1 <-  which(sips.and.scans$sips_diff_days1 > 364)
sip.post.scan2 <-  which(sips.and.scans$sips_diff_days2 > 364)
sip.post.scan3 <-  which(sips.and.scans$sips_diff_days3 > 364)

#get BBLIDs to go with the timepoints
scan1_b <- as.integer(sips.and.scans$BBLID[sip.pre.scan1])
scan2_b <- as.integer(sips.and.scans$BBLID[sip.pre.scan2])
scan3_b <- as.integer(sips.and.scans$BBLID[sip.pre.scan3])
scan1_f <- as.integer(sips.and.scans$BBLID[sip.post.scan1])
scan2_f <- as.integer(sips.and.scans$BBLID[sip.post.scan2])
scan3_f <- as.integer(sips.and.scans$BBLID[sip.post.scan3])

#label SIPS relative to each scan
sips.and.scans.labeled <- sips.and.scans %>%
  mutate(scan1.time = case_when(row_number() %in% sip.pre.scan1 ~ "base", 
                                row_number() %in% sip.post.scan1 ~ "end"),
         scan2.time = case_when(row_number() %in% sip.pre.scan2 ~ "base", 
                                row_number() %in% sip.post.scan2 ~ "end"),
         scan3.time = case_when(row_number() %in% sip.pre.scan3 ~ "base", 
                                row_number() %in% sip.post.scan3 ~ "end"),
         bblid_scan = paste(as.character(BBLID), as.character(scan_date_7t_1))) %>% #making unique ID - temporary until i pull scan ids
  filter_at(vars(scan1.time, scan2.time, scan3.time), any_vars(!is.na(.)))

#label which scan should be used by BBLID
sips.and.scans.labeled$use_scan1 <- (as.integer(sips.and.scans.labeled$BBLID) %in% scan1_b & as.integer(sips.and.scans.labeled$BBLID) %in% scan1_f)
sips.and.scans.labeled$use_scan2 <- (as.integer(sips.and.scans.labeled$BBLID) %in% scan2_b & as.integer(sips.and.scans.labeled$BBLID) %in% scan2_f)
sips.and.scans.labeled$use_scan3 <- (as.integer(sips.and.scans.labeled$BBLID) %in% scan3_b & as.integer(sips.and.scans.labeled$BBLID) %in% scan3_f)

#only keep clinical values that match a scan
sips.sops.lab_filt <- sips.and.scans.labeled %>%
  filter((use_scan1==TRUE & (!is.na(scan1.time)))|
         (use_scan2==TRUE & (!is.na(scan2.time)))| 
         (use_scan3==TRUE & (!is.na(scan3.time))))
n_unique(sips.sops.lab_filt$BBLID)

#see which scans timepoints are actually IDd for use across any subj
any(unique(sips.sops.lab_filt$use_scan1))
any(unique(sips.sops.lab_filt$use_scan2))
any(unique(sips.sops.lab_filt$use_scan3))
```

Ok, now we know which scan to use by BBLID (based on SIPS dates) and have labeled if each SIPS assessment is a baseline or endpoint for which scans.

```{r}
#pivot longer
scanlist_messy <- sips.sops.lab_filt %>% 
  select(BBLID,scan_date_7t_1,scan_scanner_1,scan_SCANID_1,use_scan1, scan_date_7t_2,scan_scanner_2,scan_SCANID_2,use_scan2, 11:48, 50, 51, 53) %>%
  distinct() %>%
  rename(date7t__1=scan_date_7t_1,
         date7t__2=scan_date_7t_2,
         scanner__1=scan_scanner_1,
         scanner__2=scan_scanner_2,
         scanid__1=scan_SCANID_1,
         scanid__2=scan_SCANID_2,
         use.scan__1=use_scan1,
         use.scan__2=use_scan2,
         sips_diff_days__1=sips_diff_days1,
         sips_diff_days__2=sips_diff_days2,
         scantime__1=scan1.time,
         scantime__2=scan2.time) %>%
    pivot_longer(!BBLID, names_to = c(".value", "scan"), names_sep = "__", values_drop_na = TRUE)

#the pivot isn't perfect (puts actual SIPS values in their own row) so we'll fill up values into correct slots then drop unnecessary rows
scanlist <- scanlist_messy %>%
  group_by(BBLID) %>%
  fill(7:42, .direction = "up") %>%
  ungroup() %>%
  subset(use.scan==TRUE) %>%
  mutate(bblid_scan = paste(as.character(BBLID), as.character(scan)))

#if ppt has multiple baseline scans, just select the one closest to scan date
scanlist_clean <- scanlist %>%
  group_by(bblid_scan, scantime) %>%
  arrange(abs(sips_diff_days)) %>%
  mutate(row = row_number()) %>%
  filter(!(scantime=="base" & row > 1)) %>% 
  drop_na(scantime) %>%
  mutate(scantime=paste(scantime,as.character(row), sep=".")) %>% #number endpoints if subj has more than one
  select(-row)%>%
  ungroup()

table(scanlist_clean$scantime)
```

Ok, now we calculate change from baseline to each endpoint (absolute, percentage, and slope). Change is defined as final score - baseline score, so a negative value means final<baseline which means sxs improved! (except GAFC, where positive -> sxs improved)

``` {r}
#calculate change
sops_change <- scanlist_clean %>% 
  group_by(bblid_scan) %>%
  arrange(scantime) %>%
  mutate(abs.delta.sops_p = sops_p - first(sops_p), #change=final-baseline -> negative value means score improves!
         abs.delta.sops_n = sops_n - first(sops_n),
         abs.delta.sops_d = sops_d - first(sops_d),
         abs.delta.sops_g = sops_g - first(sops_g),
         abs.delta.sops_tot = sops_tot - first(sops_tot),
         abs.delta.gafc = GAF_C - first(GAF_C),
         pct.delta.sops_p = abs.delta.sops_p/first(sops_p)  * 100,
         pct.delta.sops_n = abs.delta.sops_n/first(sops_n)  * 100,
         pct.delta.sops_d = abs.delta.sops_d/first(sops_d)  * 100,
         pct.delta.sops_g = abs.delta.sops_g/first(sops_g)  * 100,
         pct.delta.sops_tot = abs.delta.sops_tot/first(sops_tot)  * 100,
         pct.delta.gacf = abs.delta.gafc/first(GAF_C)  * 100,
         delta.sops_time = as.numeric(difftime(DOSIPS,first(DOSIPS), units = "days")),
         slope.sops_p = abs.delta.sops_p/delta.sops_time,
         slope.sops_n = abs.delta.sops_n/delta.sops_time,
         slope.sops_d = abs.delta.sops_d/delta.sops_time,
         slope.sops_g = abs.delta.sops_g/delta.sops_time,
         slope.sops_tot = abs.delta.sops_tot/delta.sops_time,
         slope.gafc = abs.delta.gafc/delta.sops_time,
         observation = paste(bblid_scan, scantime)) %>% #add unique ID for each BBLID-scan-endpoint#
  ungroup() %>%
  select(!c(sops.item.list, sops_p, sops_n, sops_d, sops_g, sops_tot)) %>% #drop raw scores
  filter(!(scantime=="base.1"))

#double check to make sure no endpoint scans come before baseline scans 
summary(sops_change$delta.sops_time)
#just one row per BBLID-scan-endpoint#
any(duplicated(sops_change$observation))
n_unique(sops_change$observation)

#fun plots
ggplot(sops_change, aes(x=slope.sops_tot)) + geom_histogram()
ggplot(sops_change, aes(x=slope.sops_n)) + geom_histogram()
ggplot(sops_change, aes(x=slope.gafc)) + geom_histogram()

scanlist_clean %>%
ggplot(aes(x = sips_diff_days, y = sops_tot, color = bblid_scan)) + 
    geom_line() + 
    theme_bw()
```
linear model for fun

```{r}
bs.fit <- lm(slope.sops_tot ~ PRODROMAL, data=sops_change)
Anova(bs.fit)
```

# Adding Covariates

## Demographics
Adding sex, DOB/age, race (csv adapted from 7T list from David).

```{r}
#load demographics (adapted from 7T list from David)
demo <- read.csv("data/7t_demographics.csv", header = TRUE, stringsAsFactors = TRUE, na = "") %>%
  mutate(sex = as.factor(Sex_1M),
         Ethnicity = as.factor(Ethnicity),
         DOB = as.Date(DOB, format = "%m/%d/%y"),
         BBLID = as.factor(BBLID)) %>%
  select(-Diagnostic_Group, -sex, -Age)

sops_change_demo <- left_join(sops_change, demo, by="BBLID")
skim(sops_change_demo)
n_unique(sops_change_demo$observation)
```

Demographic variables have 100% completion

## Diagnosis

Now merge in diagnosis and filter to smallest datediff

```{r}
sops_change_dx <- left_join(sops_change_demo, redcap, by="BBLID")
any(is.na(sops_change_dx$AXIS1_DX1)) #no primary dx missing

#calculate datediff from dx to scan
sops_change_dx <- sops_change_dx %>%
  mutate(dx_7t_diff = as.numeric(difftime(DODIAGNOSIS,date7t, units = "days"))) %>%
  arrange(abs(dx_7t_diff)) %>%
  group_by(observation) %>%
  slice_head() %>% #keep only closest dx
  ungroup() %>%
  select_if(~!all(is.na(.))) #drop any col that's totally empty

#make sure no obs lost
n_unique(sops_change_dx$observation)
summary(sops_change_dx$dx_7t_diff)
#skim(sops_change_dx)
```

Just keeping diagnosis at baseline for now! It's probable that some dx changed over the course of the study but not what we're interested in ATM. 

linear model for fun

```{r}
bs.fit2 <- lm(slope.sops_tot ~ AXIS1_DX1, data=sops_change_dx)
Anova(bs.fit2)
```

## Medication Info

Load and clean med data

```{r}
#load med data
meds <- read.csv("data/oracle_meds_all.csv", header = TRUE, stringsAsFactors = F, na = c("", "null"))
skim(meds)

#each med is it's own row, so need to rotate wide
meds <- meds %>%
  group_by(BBLID, DOCOLLECT) %>%
  mutate(ID = paste(as.character(BBLID), as.character(DOCOLLECT)), #col to ID subj+timepoint
         count = row_number()) %>% #count meds entered
  ungroup()

hist(meds$count)

medswide <- meds %>%
  select(BBLID, DOCOLLECT, ID, count, MEDICINE, NOTES, IS_CURRENT) %>%
  mutate(BBLID=as.factor(BBLID)) %>%
  pivot_wider(., names_from = count, values_from=c(MEDICINE, NOTES, IS_CURRENT)) # pivot
```

Merge and keep med info at time of scan as well as time of each SOPS endpoint assessment. We'll restrict the meds to be within 

```{r}
sops_change_meds <- left_join(sops_change_dx, medswide, by="BBLID")

#calculate datediff from meds to scan/SIPS
sops_change_meds <- sops_change_meds %>%
  mutate(meds_7t_diff = as.numeric(difftime(DOCOLLECT, date7t, units = "days")),
         meds_sops_diff = as.numeric(difftime(DOCOLLECT, DOSIPS, units = "days")))
#ok, now pull the baseline meds and set them aside for now
meds_baseline <- sops_change_meds %>%
  arrange(abs(meds_7t_diff)) %>%
  group_by(observation) %>%
  slice_head() %>% #keep only closest dx
  ungroup() %>%
  select(observation, meds_7t_diff, DOCOLLECT | contains("MEDICINE_") | contains("NOTES_") | contains("IS_CURRENT_")) %>% #drop non-med stuff
  select_if(~!all(is.na(.))) %>% #drop any col that's totally empty
  rename_with(.cols = 3:ncol(.), function(x){paste0("base.", x)}) #add prefix
#go back and filter down to endpoint meds
meds_end <- sops_change_meds %>%
  arrange(abs(meds_sops_diff)) %>%
  group_by(observation) %>%
  slice_head() %>% #keep only closest dx
  ungroup() %>%
  select(observation, meds_sops_diff, DOCOLLECT | contains("MEDICINE_") | contains("NOTES_") | contains("IS_CURRENT_")) %>% #drop non-med stuff
  select_if(~!all(is.na(.)))

#plots to see the range of closest med info
sops.med.p <- meds_end %>%
  ggplot() + 
    geom_histogram(aes(x = meds_sops_diff), position="identity", color="lightblue", fill="lightblue")
scan.med.p <- meds_baseline %>%
  ggplot() + 
    geom_histogram(aes(x = meds_7t_diff), position="identity", color="lightgreen", fill="lightgreen")
# put p1 and p2 together
ggarrange(sops.med.p, scan.med.p, common.legend = T, legend = "bottom", ncol = 2)
```

Ok, maybe let's restrict to +- 100 days and add back in. Also tried 200 days but didn't add anything

```{r}
#filter down
meds_baseline <- meds_baseline %>%
  filter(abs(meds_7t_diff) <= 100) %>%
  mutate(across(c(contains("MEDICINE_")), factor)) #also setting meds as factors to see frequency later
meds_end <- meds_end %>%
  filter(abs(meds_sops_diff) <= 100) %>%
  mutate(across(c(contains("MEDICINE_")), factor)) #also setting meds as factors to see frequency later


#merge base and end meds together
meds_filt <- full_join(meds_baseline, meds_end, by="observation")
sum(is.na(meds_filt$meds_7t_diff))
sum(is.na(meds_filt$meds_sops_diff))

#merging meds with sops_change_dx b/c sops_change_meds already has med stuff
sops_change_meds.final <- left_join(sops_change_dx, meds_filt, by="observation") 

#make sure no obs lost
n_unique(sops_change_meds.final$observation)
#skim(sops_change_meds.final)
```

Ok, now how to summarize all this medication info into one cohesive value

```{r}
#see how often each med is reported
sops_change_meds.final %>%
  select(contains("MEDICINE")) %>%
  unlist() %>%
  table()
```

NOTES:
CARVEDILOL = beta blocker (anxiety?)
CLONIDINE = hypertension, ADHD?
DEXMETHYLPHENIDATE HYDROCHLORIDE = ADHD
NORETHINDRONE ACETATE = progestin
RANITIDINE = antihistamine/antacid
TRILOMARZIA = birth control
ALBUTEROL = asthma
DEXEDRINE = ADHD
DICLOFENAC = arthritis
ADVAIR = asthma

ideas - use 0,1 indicator for each class of medication at each timepoint? could be granular (asthma, allergies, stimulants) or more broad (psych_rx, other_rx)
- use that indicator but lifetime?

keep if start/end dates of med are < date > ?
lifetime indicator - dont filter by date assessed, just if date-rx-start is < endpoint date?

Test using the DOMED_START and DOMED_END to only capture meds that are explicitly current for 7T scan

```{r}
#get subset
meds_reduced <- meds %>%
  select(BBLID, DOCOLLECT, MED_COLLECT, NOTES, MEDICINE, PHARMCLASS, DOMED_START, DOMED_END, IS_CURRENT)

#get managable subset of scan list
scan_dates <- sops_change_dx %>%
  select(BBLID, date7t, observation) %>%
  distinct()

#merge
scan_date_med <- left_join(scan_dates, meds_reduced, by="BBLID")

#ok, keep only rows where med is current to 7T
scan_date_med_current <- scan_date_med %>%
  filter(is.na(DOMED_START) | date7t > DOMED_START) %>% #keep if med is STARTED BEFORE 7t
  filter(is.na(DOMED_START) | date7t < DOMED_END) %>% #keep if med was ENDED AFTER 7t
  mutate(med_7t_diff = as.numeric(difftime(DOCOLLECT,date7t, units = "days"))) %>%
  filter(!(is.na(DOMED_START) & (abs(med_7t_diff) > 100))) %>%  #only keep NAs if DOCOLLECT is w/in 100 days of scan
  group_by(observation) %>%
  mutate(count = row_number()) %>% #add counts so i can pivot after
  ungroup() %>%
  rename_with(.cols = 4:ncol(.), function(x){paste0("base.", x)}) #add prefix
scan_date_med_current$base.NOTES[is.na(scan_date_med_current$base.NOTES)] <- "empty entry"
skim(scan_date_med_current)

#pivot wide
scan_meds_current_wide <- pivot_wider(scan_date_med_current, id_cols=observation, names_from = base.count, values_from=c(4:12))
#just keep essential cols
scan.meds.current_wide.filt <- scan_meds_current_wide %>%
  select(observation | contains("MEDICINE_") | contains("NOTES_")) %>%
  select_if(~!all(is.na(.))) #drop any col that's totally empty
```

Trying to add PNC data to see if that will help get more baseline info

```{r}
pnc.meds <- read.csv("data/n1601_health_with_meds_20170421.csv", header = TRUE, stringsAsFactors = F, na ="") %>%
  mutate(BBLID=as.factor(bblid)) %>%
  select(-bblid)
skim(pnc.meds)

pnc.meds_scan_date <- inner_join(scan_dates, pnc.meds, by="BBLID") %>%
  distinct(BBLID, date7t, .keep_all = T)
n_unique(pnc.meds_scan_date$BBLID)
```

Waiting on Kosha to get the dates, hold off for now. In the meantime I'll try to find meds current to endpoint SOPs.

```{r}
#get managable subset of endpoint list
end_dates <- sops_change_dx %>%
  select(BBLID, DOSIPS, observation) %>%
  distinct()
#merge
end_date_med <- left_join(end_dates, meds_reduced, by="BBLID")

#ok, keep only rows where med is current to 7T
end_date_med_current <- end_date_med %>%
  filter(is.na(DOMED_START) | DOSIPS > DOMED_START) %>% #keep if med is STARTED BEFORE SOPS end
  filter(is.na(DOMED_START) | DOSIPS < DOMED_END) %>% #keep if med was ENDED AFTER SOPS end
  mutate(med_sops_diff = as.numeric(difftime(DOCOLLECT, DOSIPS, units = "days"))) %>%
  filter(!(is.na(DOMED_START) & (abs(med_sops_diff) > 100))) %>%  #only keep NAs if DOCOLLECT is w/in 100 days of scan
  group_by(observation) %>%
  mutate(count = row_number()) %>% #add counts so i can pivot after
  ungroup()
end_date_med_current$NOTES[is.na(end_date_med_current$NOTES)] <- "empty entry"
skim(end_date_med_current)

#pivot wide
end_meds_current_wide <- pivot_wider(end_date_med_current, id_cols=observation, names_from = count, values_from=c(4:12))
#just keep essential cols
end.meds.current_wide.filt <- end_meds_current_wide %>%
  select(observation | contains("MEDICINE_") | contains("NOTES_")) %>%
  select_if(~!all(is.na(.))) #drop any col that's totally empty
```

This seems to be working, lets try merging baseline and endpoint meds and see what we have

```{r}
currentmeds <- right_join(scan.meds.current_wide.filt, end.meds.current_wide.filt, by="observation")
```

Actually very narrow range of meds here, but the `empty entry` notes indicate either no medications or no new medications. Therefore I'll need to look back at participant's former meds

Lets go back and look at lifetime:

```{r}
#ok, keep only rows where med is current to 7T
med_lifetime <- end_date_med %>%
  filter(is.na(DOMED_START) | DOSIPS > DOMED_START) %>% #keep if med is STARTED BEFORE SOPS end
  group_by(observation) %>%
  mutate(count = row_number()) %>% #add counts so i can pivot after
  ungroup()
med_lifetime$MEDICINE[is.na(med_lifetime$MEDICINE)] <- "empty entry"
skim(med_lifetime)

#trying to find distinct meds for each observation
med_lifetime_dist <- med_lifetime %>%
  group_by(observation) %>%
  distinct(MEDICINE, .keep_all = T) %>% #keep distinct
  mutate(count=row_number()) %>% #update counts so i can pivot after
  ungroup() %>%
  select(BBLID, DOSIPS, observation, MEDICINE, NOTES, count)
table(med_lifetime_dist$MEDICINE)

#pivot
med_lifetime_dist.wide <- pivot_wider(med_lifetime_dist, id_cols=observation, names_from = count, values_from=c(MEDICINE, NOTES)) %>% # pivot
  select_if(~!all(is.na(.))) #drop any col that's totally empty
n_unique(med_lifetime_dist.wide$observation)
sum(is.na(med_lifetime_dist.wide$MEDICINE_1))
```

Because there's med info for all subjects, if an individual only has `empty entry` value then we can infer that they are not taking any medications.

# CEST values -> FINAL DF

# Modeling

## SIPS/SOPS as Change


