---
title: "roalf_longitudinal"
output: html_document
---

A cleaner/more streamlined version of `rotation_exploration.Rmd` for assessing changes in clinical scores pre-post 7T glucest.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load packages

```{r Install/Load packages, echo=FALSE, message=FALSE, warning=FALSE}
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(tidyverse, ggplot2, ggpubr, ggrepel, skimr, lubridate, data.table)
```

## 7T dates

See `scan_list_comp.Rmd`.

``` {r load.7t, echo=FALSE}
load("data/all_7T_april22.Rdata")
all_7t.wide$BBLID <- as.integer(all_7t.wide$BBLID)
```

## Load All Data

*Notes* 
* Per David, no PNC subj should be 22q - no need to worry about filtering beyond those ID'd by RedCap
* no date available for PNC dx, arbitrarily picking Jan 1 2000

*TO DO* 
+ get actual PNC clinical dates
+ filter SYRP datasets to keep pre and/or self-report when subj has multiple proband assessments on the same day
+ get medication info

```{r load.clinical, echo=FALSE}
#load PNC data
goassess <- read.csv("/Users/megardn/ImageData/PMACS_remote/clinical/goassess_diagnosis.csv", header=T) %>%
  mutate(dx = as.factor(goassessDxpmr6),
        dx_source = "goassess",
        dx_date = as.Date("2000-01-01")) %>% #no date available for PNC dx, arbitrarily picking Jan 1 2000
        select (bblid, dx, dx_source, dx_date)

#load RedCap diagnosis data
redcap <- read.csv("data/diagnosis_long.csv", header = TRUE, stringsAsFactors = F, na = "") %>% 
  select(BBLID, DODIAGNOSIS, HSTATUS, AXIS1_DX1, allaxis1desc, GASR_CURRENT, AGEONSET_AXIS1, AGEONSET_CLINICRISK, dx_pscat, timepoint, FROM_CAPA, FROM_DIGS, FROM_SCID, FROM_FIGS, FROM_RECORDS, FROM_MINI, FROM_SIPS) %>% #get manageable subset of data
  mutate(DODIAGNOSIS = as.Date(DODIAGNOSIS))

#load SIPS
sips <- fread("data/oracle_sips_all.csv", na.strings="null", header=TRUE)
  #mutate(DOSIPS = as.Date(DOSIPS),
         #bblid = as.factor(bblid))
sips$BBLID <- as.factor(sips$BBLID)
# skim(pnc.sips)
# skim(sips)

#load PRIME
prime <- read.csv("data/prime_long.csv", header = TRUE, stringsAsFactors = F, na = "") %>%
  mutate(date = as.Date(date))
pnc.prime <- read.csv("/Users/megardn/ImageData/PMACS_remote/clinical/n1601_goassess_sips.csv", header=T)

#load CGAS
cgas <- read.csv("data/c-gas.csv", header = TRUE, stringsAsFactors = F, na = "") %>%
  mutate(date = as.Date(date, format = "%m/%d/%y %H:%M"))

#load SYRPs
syrp_roalf <- read.csv("data/7TSharedScalesCollec-Roalf7TProjects_DATA_2022-02-11_1858.csv", header = TRUE, stringsAsFactors = F, na = "")
syrp_sat <- read.csv("data/7TSharedScalesCollec-Satterthwaite7T_DATA_2022-02-11_1858.csv", header = TRUE, stringsAsFactors = F, na = "")
syrp <- rbind(syrp_roalf, syrp_sat) %>% 
  mutate(date = as.Date(date),
         bblid = as.factor(bblid))
table(duplicated(syrp)) #no duplicate rows across Roalf and Satterthwiate csvs

#load demographics (adapted from 7T list from David)
demo <- read.csv("data/7t_demographics.csv", header = TRUE, stringsAsFactors = TRUE, na = "") %>%
  mutate(sex = as.factor(Sex_1M),
         ethnicity = as.factor(Ethnicity),
         DOB = as.Date(DOB, format = "%m/%d/%y"),
         BBLID = as.factor(BBLID))
```


``` {r 22q, echo=FALSE}
#make list of 22q subj
q22 <- subset(redcap, HSTATUS == "22q", select = BBLID) %>% unique()

#remove 22q participants from scan list, that way it'll apply to all subsequent merges w/ clinical data
all_7t.wide <- all_7t.wide %>% subset(!(BBLID %in% q22$BBLID)) #drop 22q sub
length(unique(all_7t.wide$BBLID))
```

There are `r length(unique(all_7t.wide$BBLID))` (132) subjects who have 7T scans, excluding 22q pts ID'd from Kosha's RedCap data

## Diagnoses & Demographics

Won't be the focus of our analyses but will be useful for grouping. Merging diagnoses from PNC and Kosha's RedCap pull:

``` {r dx.clean}
#first clean redcap dx source info
names(redcap)
sources <- c("FROM_CAPA", "FROM_DIGS", "FROM_SCID", "FROM_FIGS", "FROM_RECORDS", "FROM_MINI", "FROM_SIPS")
redcap[,sources] <- redcap[,sources] %>% 
  transmute_all(funs(ifelse(. == 1, deparse(substitute(.)), NA))) #replace coding with col name
redcap <- redcap %>% unite(dx_source, c(sources), na.rm = TRUE, remove=FALSE) #merge across - not pretty but gets the point across for now

redcap_dx <- redcap %>%
  rename(bblid=BBLID, dx_date=DODIAGNOSIS, dx=dx_pscat) %>%
  mutate(dx_date=as.Date(dx_date)) %>%
  select (bblid, dx, dx_source, dx_date)
  

all_dx <- rbind(redcap_dx,goassess)%>% 
  subset(!(bblid %in% q22$BBLID)) #drop 22q sub
id_date <- all_dx %>% select(bblid, dx_date)
table(duplicated(id_date))#no duplicate rows, all ID and dx_date combos unique!
```

Now combine dx with scan dates:

``` {r dx.match}
dx_scan_date <- right_join(all_dx, all_7t.wide, by = c("bblid" = "BBLID")) %>%
  mutate(bblid = as.factor(bblid))

length(unique(dx_scan_date$bblid))
sum(is.na(dx_scan_date$dx)) #some subjects with scans have no diagnosis from Kosha's redcap or goassess

dx_missing <- dx_scan_date %>% subset(is.na(dx), select = c(bblid, scan_7T_date_1))
# write.csv(dx_missing, "missing_dx.csv", row.names=FALSE)
```

There are `r sum(is.na(dx_scan_date$dx))` (67!) subj who have 7T scans but are missing diagnoses from redcap/PNC!

Seem to have some data for these ppts in spreadsheets David added/demographics from 7T list - reviewed with David. *TO DO* - get newer dx from Arianna and Heather 

Proceeding for now with dx available, recoding factors for consistency between PNC and RedCap when possible:

``` {r dx.factor}
table(dx_scan_date$dx)
dx_scan_date <- dx_scan_date %>% mutate(dx = as.factor(dx))
dx_scan_date$dx <- fct_collapse(dx_scan_date$dx, "healthy" = c("noDSMdx", "TD"),
                                "other_dx" = c("OP", "other"),
                                "psych_spect" = c("psy", "PS", "pro")) #recode factors

#make new df with only baseline dx (closest to 7T) adding dx Arianna pulled from oracle
#calc diff (negative values -> dx before 7t)
dx_scan_date$dx_diff = as.numeric(difftime(dx_scan_date$dx_date, dx_scan_date$scan_7T_date_1, units = "days"))
#select one dx per subject with date closest to &t
dx_baseline <- dx_scan_date %>%
  mutate(abs_diff_in_days=abs(dx_diff)) %>%
  group_by(bblid) %>%
  arrange(abs_diff_in_days) %>%
  slice(c(1)) %>%
  ungroup() %>%
  select(!c(abs_diff_in_days))

summary(dx_baseline$dx_diff) #not great alignment between date of dx and date of scan
ggplot(dx_baseline, aes(x=dx_diff, y=bblid, color=dx)) + 
  geom_point() +
  labs(title="Dx Relative to First 7T Scan", x="Days from 7T Scan") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

Adding in demographics - will have dx from 7T list in separate column for now, since I'm  not sure when/where they came from

``` {r demo}
#recode & rename variables for clarity
demo <- demo %>% select(-Sex_1M, -Ethnicity, -Age)
demo$sex <- recode(demo$sex, "1" = "M", "2" = "F")

#merge with first 7t list
dx_baseline <- left_join(dx_baseline, demo, by = c("bblid"="BBLID"))

#calculate age at first 7t
dx_baseline$age_scan = as.period(interval(start = dx_baseline$DOB, end = dx_baseline$scan_7T_date_1))$year
dx_baseline <- dx_baseline %>% select(-DOB) #drop DOB since we don't need that PHI anymore
```

## SIPS & SOPS

Some subjects have multiple assessments on one day (one from proband one from family, etc); need to be sure to pick only `TYPE == "COMBINED"` per discussion with David. If no combined, pick proband.

``` {r sips&sops}
#SIPS + first 7t and dx
sips_date <- inner_join(sips, dx_baseline, by = c("BBLID"="bblid"))
length(unique(sips_date$BBLID))
#calc diff (negative values -> dx before 7t)
sips_date$diff_in_days1 = as.numeric(difftime(sips_date$DOSIPS, sips_date$scan_7T_date_1, units = "days"))
sips_date$diff_in_days2 = as.numeric(difftime(sips_date$DOSIPS, sips_date$scan_7T_date_2, units = "days"))
sips_date$diff_in_days3 = as.numeric(difftime(sips_date$DOSIPS, sips_date$scan_7T_date_3, units = "days"))
#str(sips_date$diff_in_days3)
# drop data more than 1 yr prior to first scan
sips_date <- subset(sips_date, diff_in_days1 >= -365)
length(unique(sips_date$BBLID)) #no subj lost
min(sips_date$diff_in_days1) #confirm

#list sops items
plist <- c("P1", "P2", "P3", "P4", "P5")
nlist <- c("N1", "N2", "N3", "N4", "N5", "N6")
dlist <- c("D1", "D2", "D3", "D4")
glist <- c("G1", "G2", "G3", "G4")
sops.item.list <- c(plist, nlist, dlist, glist)

#drop missing sops items
sips_date.filt <- sips_date %>% drop_na(c(sops.item.list))
#sum(is.na(sips_date.filt[,c(sops.item.list)]))

#if subj has more than one assessment for same day, select Combined 
sips.sops_scan <- sips_date.filt %>% mutate(TYPE = as.factor(TYPE))
levels(sips.sops_scan$TYPE) 
sips.sops_scan$TYPE <- fct_collapse(sips.sops_scan$TYPE, "Proband" = c("Proband", "proband", "P", "FP", "IP"), "Combined"=c("Combined", "FPC")) #combine all proband
sips.sops_scan$TYPE <- factor(sips.sops_scan$TYPE, levels = c("Combined", "Proband", "Collateral"))

#group data to select correct type (defined as lowest ordered level) - also cleaning up
sips.sops_scan.filt <- sips.sops_scan %>%
  group_by(BBLID, DOSIPS) %>%
  arrange(TYPE) %>%
  slice(c(1)) %>%
  ungroup() %>%
  select(-SIPS_SOURCEID, -SOURCE_PROJECT) %>%
  mutate(DOSIPS=as.Date(DOSIPS))

length(unique(sips.sops_scan.filt$BBLID)) #confirm no subjects deleted
skimr::skim(sips.sops_scan.filt)
head(sips.sops_scan.filt)
```

Plotting all SIPS clin vs all 7T scans

```{r}
# create base scatter plot
plot(sips.sops_scan.filt$DOSIPS, sips.sops_scan.filt$BBLID, col='black')
  
# overlay scatter plot 
points(sips.sops_scan.filt$scan_7T_date_1, sips.sops_scan.filt$BBLID, col='green', pch=12)
points(sips.sops_scan.filt$scan_7T_date_2, sips.sops_scan.filt$BBLID, col='red', pch=13)
points(sips.sops_scan.filt$scan_7T_date_3, sips.sops_scan.filt$BBLID, col='blue')
```

```{r}
# #label SIPS relative to each scan
# sips.sops_scan.filt$scan1_tp <- findInterval(sips.sops_scan.filt$diff_in_days1, c(-365, 364))
# sips.sops_scan.filt$scan2_tp <- findInterval(sips.sops_scan.filt$diff_in_days2, c(-365, 364))
# sips.sops_scan.filt$scan3_tp <- findInterval(sips.sops_scan.filt$diff_in_days3, c(-365, 364))
# 
# clin364_sip1 <-  sips.sops_scan.filt  %>% subset(diff_in_days1 <= 364, select = BBLID) %>% unique()
# ##find subjects that have a followup sips sops >365 days post-7T 
# clin365_sip1 <- sips.sops_scan.filt %>% subset(diff_in_days1 >= 365, select = BBLID) %>% unique()

sip.pre.scan1 <-  which(sips.sops_scan.filt$diff_in_days1 <= 364)
sip.pre.scan2 <-  which(sips.sops_scan.filt$diff_in_days2 <= 364 & sips.sops_scan.filt$diff_in_days2 >= -365)
sip.pre.scan3 <-  which(sips.sops_scan.filt$diff_in_days3 <= 364 & sips.sops_scan.filt$diff_in_days3 >= -365)
#find subjects that have a followup sips sops >365 days post-7T 
sip.post.scan1 <-  which(sips.sops_scan.filt$diff_in_days1 > 364)
sip.post.scan2 <-  which(sips.sops_scan.filt$diff_in_days2 > 364)
sip.post.scan3 <-  which(sips.sops_scan.filt$diff_in_days3 > 364)

#labe

genes %>%
mutate(Sig = ifelse(row_number() %in% sig, "YES", "NO"))

scan1_b <- as.integer(sips.sops_scan.filt$BBLID[sip.pre.scan1])
scan2_b <- as.integer(sips.sops_scan.filt$BBLID[sip.pre.scan2])
scan3_b <- as.integer(sips.sops_scan.filt$BBLID[sip.pre.scan3])
scan1_f <- as.integer(sips.sops_scan.filt$BBLID[sip.post.scan1])
scan2_f <- as.integer(sips.sops_scan.filt$BBLID[sip.post.scan2])
scan3_f <- as.integer(sips.sops_scan.filt$BBLID[sip.post.scan3])

sips.sops_scan.filt$use_scan1 <- (as.integer(sips.sops_scan.filt$BBLID) %in% scan1_b & as.integer(sips.sops_scan.filt$BBLID) %in% scan1_f)
sips.sops_scan.filt$use_scan2 <- (as.integer(sips.sops_scan.filt$BBLID) %in% scan2_b & as.integer(sips.sops_scan.filt$BBLID) %in% scan2_f)
sips.sops_scan.filt$use_scan3 <- (as.integer(sips.sops_scan.filt$BBLID) %in% scan3_b & as.integer(sips.sops_scan.filt$BBLID) %in% scan3_f)

sips.sops_stuff <- subset(sips.sops_scan.filt, (use_scan2==TRUE | use_scan1==TRUE |use_scan3==TRUE)) #basically follow-up scans add no participants :(

# sorting <- sips.sops_scan.filt %>% select(BBLID,DOSIPS,scan1_tp,scan2_tp,scan3_tp) %>%
#   group_by(BBLID) %>%
#   summary()
# 
# 
# #find subjects that have a baseline sips and sops up to 364 days post-7T for each scan
# clin364_sip1 <-  sips.sops_scan.filt  %>% subset(diff_in_days1 <= 364, select = BBLID) %>% unique()
# ##find subjects that have a followup sips sops >365 days post-7T 
# clin365_sip1 <- sips.sops_scan.filt %>% subset(diff_in_days1 >= 365, select = BBLID) %>% unique()
# sips.sops_scan.scan1 <- subset(sips.sops_scan.filt, (BBLID %in% clin365_sip1$BBLID & BBLID %in% clin364_sip1$BBLID))
# n_unique(sips.sops_scan.scan1$BBLID)
# 
# clin364_sip1 <-  which(sips.sops_scan.filt$diff_in_days1 <= 364)
# clin364_sip2 <-  which(sips.sops_scan.filt$diff_in_days2 <= 364 & sips.sops_scan.filt$diff_in_days2 >= -365)
# clin364_sip3 <-  which(sips.sops_scan.filt$diff_in_days3 <= 364 & sips.sops_scan.filt$diff_in_days3 >= -365)
# #find subjects that have a followup sips sops >365 days post-7T 
# clin365_sip1 <-  which(sips.sops_scan.filt$diff_in_days1 > 364)
# clin365_sip2 <-  which(sips.sops_scan.filt$diff_in_days2 > 364)
# clin365_sip3 <-  which(sips.sops_scan.filt$diff_in_days3 > 364)
# #scan1_f <- as.integer(sips.sops_scan.filt$BBLID[clin365_sip1])
# scan2_f <- as.integer(sips.sops_scan.filt$BBLID[clin365_sip2])
# scan3_f <- as.integer(sips.sops_scan.filt$BBLID[clin365_sip3])
# 
# sips.sops_scan.scan1 <- subset(sips.sops_scan.filt, (BBLID %in% clin365_sip1$BBLID & BBLID %in% clin364_sip1$BBLID))
# n_unique(sips.sops_scan.scan1$BBLID)
# 
# #filter df on date
# sips.sops_scan.filt <- subset(sips.sops_scan.filt, (BBLID %in% clin365_sip$BBLID & BBLID %in% clin364_sip$BBLID))

length(unique(sips.sops_stuff$BBLID))

#plot for fun
ggplot(sips.sops_stuff, aes(x=diff_in_days1, y=BBLID)) + 
  geom_point() +
  geom_vline(xintercept=365) +
  labs(title="SOPS Relative to First 7T Scan", x="Days from 7T Scan") +
  theme(legend.position = "none")

skim(sips.sops_stuff)
```

Note - adding follow-up scans doesn't actually give more subjects, but i'll plan on processing scan1 and 2 for the subj who have `use_scan2==TRUE` just in case there are data quality problems with scan 1.

analysis ideas from jake: 
- model linear regression for sops score for each subject, predict slope
- mixed effects model (can incorporate all clinical timepoints, subjects dont have to have same number)


```{r}
#return scanlist
scanlist <- sips.sops_stuff %>% select(BBLID,scan_7T_date_1,scan_scanner_1,scan_SCANID_1,use_scan1, scan_7T_date_2,scan_scanner_2,scan_SCANID_2,use_scan2) %>%
  distinct() %>%
  rename(date7t_1=scan_7T_date_1,
         date7t_2=scan_7T_date_2,
         scanner_1=scan_scanner_1,
         scanner_2=scan_scanner_2,
         scanid_1=scan_SCANID_1,
         scanid_2=scan_SCANID_2,
         use.scan_1=use_scan1,
         use.scan_2=use_scan2) %>%
  pivot_longer(!BBLID, names_to = c(".value", "scan"), names_sep = "_", values_drop_na = TRUE) %>%
  subset(use.scan==TRUE)

write_csv(scanlist, file = "data/SCANS-TO-ANALYZE.csv")
```


### SOPS

SOPS seems to be the 12 item components of SIPS (which are overall ratings of severity).
Focusing on SOPS (summing within and subscales Pos, Neg, Gen, Disorg) - seem to be valid scoring based on ct.gov. 

``` {r sops}
#sum(is.na(sips.sops_scan.filt[7:25]))#see missing sops scores
sops_scan <- sips.sops_stuff %>% mutate(sops_p = rowSums(across(plist), na.rm=F),
                                  sops_n = rowSums(across(nlist), na.rm=F),
                                  sops_d = rowSums(across(dlist), na.rm=F),
                                  sops_g = rowSums(across(glist), na.rm=F),
                                  sops_tot = rowSums(across(c(sops_p, sops_n, sops_d, sops_g), na.rm=F)))

#sum(is.na(sops_scan$sops_tot)) #none missing!

#plot for fun
ggplot(sops_scan, aes(x=diff_in_days1, y=BBLID, color=sops_tot)) + 
  geom_point() +
  geom_vline(xintercept=365) +
  labs(title="SOPS Relative to First 7T Scan", x="Days from 7T Scan") +
  theme(legend.position = "none")

length(unique(sops_scan$BBLID))
```


Looking just at change immediately pre-post 7T:

``` {r sops.change.prox, echo=FALSE}
#define baseline - scan closest to 7T
#find smallest datediff that's >=365
post_sops <- sops_tot %>%
  select(1:21, 32:37) %>% #filter so we're not duplicating rows that wont change (demographics, 7t date, etc.)
  group_by(bblid) %>%
  filter(diff_in_days >= 365) %>%
  arrange(bblid, diff_in_days) %>%
  slice(c(1))%>% 
  rename_at(3:27, ~ paste0(.x, "_post")) #add suffix to all

#find datediff that's within 365 days and closest to 0
pre_sops <- sops_tot %>%
  group_by(bblid) %>%
  filter(diff_in_days < 365) %>%
  mutate(abs_diff_in_days=abs(diff_in_days)) %>%
  arrange(bblid, abs_diff_in_days) %>%
  slice(c(1)) %>% 
  ungroup() %>%
  select(!c(abs_diff_in_days)) %>% 
  rename_at(2:21, ~paste0(., "_pre")) %>% #add suffix to items
  rename_at(32:37, ~paste0(., "_pre")) #add suffix to scores

#new dataframe with pre-post and calculate change
proximal_sops <- inner_join(pre_sops, post_sops, by = c("bblid" = "bblid")) %>%
  mutate(change_pos = (sops_p_post - sops_p_pre),
         change_neg = (sops_n_post - sops_n_pre),
         change_dis = (sops_d_post - sops_d_pre),
         change_gen = (sops_g_post - sops_g_pre),
         change_tot = (sops_tot_post - sops_tot_pre))

#plot change
p <- ggplot(proximal_sops, aes(x=change_pos)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Positive")

n <- ggplot(proximal_sops, aes(x=change_neg)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Negative")

d <- proximal_sops %>% drop_na(change_dis) %>% 
  ggplot(aes(x=change_dis)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Disorganized")

g <- ggplot(proximal_sops, aes(x=change_gen)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="General")

sops_sub_plot <- ggpubr::ggarrange(p, n, d, g, common.legend = F, ncol = 4)
annotate_figure(p=sops_sub_plot, top = text_grob("Change in SOPS Subscales Immediately Pre-Post 7T", 
                               color = "black", face = "bold", size = 14))

ggplot(proximal_sops, aes(x=change_tot)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Change in SOPS Total Pre-Post 7T")

ggplot(proximal_sops, aes(x=change_tot, y=sops_tot_pre, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid)) +
  labs(title="Change in SOPS Total Scores Over Time", x="Change (Post - Pre)", y="Baseline SOPS Total")
```

There are `r length(unique(sops_scan$bblid))` (27) subjects with SOPS at baseline and at least 1 post-1 yr followup, though total pre-post scores are available only for `r length(unique(proximal_sops$bblid))` subjects.
Negative change -> ppt's score is getting lower (i.e. symptoms improving)

``` {r sops.prox, echo=FALSE}
proximal_sops %>% select(1, 22, 37, 63:68) %>% summary()
```

Line plot of all SOPS Total for each subj:

``` {r sops.longitudinal, echo=FALSE}
ggplot(sops_scan, aes(x=diff_in_days, y=sops_tot, color=bblid)) +
  geom_line() +
  geom_vline(xintercept=365) +
  labs(title="SOPS Total Scores Over Time")
```


### SIPS

Pulling "SIPS" ratings number of Pos, Neg, Gen, etc items on SOPS coded as 3,4,5, or 6 depending on suffix - calculating for subj that aren't already input. Not planning to use these in final analyses, but already wrote the code.

``` {r sips}
# #filter df for SOPS item completion
# sips_scan <- dx_sips.sops_scan %>% 
#   drop_na(POS_345, POS_6, NEG_3456, GEN_3456) %>% 
#   mutate(POS_3456= rowSums(across(c(POS_345, POS_6)))) #concatenating across frequency of p sxs for consistency across subscales

sips_calc <- sips.sops_scan.filt %>%
  drop_na(8:26)
#calc sips
sips_calc$sips_pos <- apply(sips_calc[8:12], 1, function(x) length(which(x>2)))
sips_calc$sips_neg <- apply(sips_calc[13:18], 1, function(x) length(which(x>2)))
sips_calc$sips_dis <- apply(sips_calc[19:22], 1, function(x) length(which(x>2)))
sips_calc$sips_gen <- apply(sips_calc[23:26], 1, function(x) length(which(x>2)))
sips_scan <- sips_calc %>% select(1, 7, 47:61)

#since dropping NA we need to refilter pre-post subj
##find subjects that have a baseline sops up to 364 days post-7T 
clin364_sip2 <-  sips_scan  %>% subset(diff_in_days <= 364, select = bblid) %>% unique()
##find subjects that have a followup sips sops >365 days post-7T 
clin365_sip2 <- sips_scan %>% subset(diff_in_days >= 365, select = bblid) %>% unique()
##filter df
sips_scan <- subset(sips_scan, (bblid %in% clin364_sip2$bblid & bblid %in% clin365_sip2$bblid))

length(unique(sips_scan$bblid))
```

There are `r length(unique(sips_scan$bblid))` (26) subjects with SIPS at baseline and at least 1 post-1 yr followup

Calculate change across time (i.e. change from one timepoint to the next):

``` {r sips.change}
#calculate change in SIPS scores, save and plot
sips_change <- sips_scan %>% 
  group_by(bblid) %>% 
  arrange(diff_in_days) %>%
  mutate("pos_diff"= sips_pos - lag(sips_pos),
         "neg_diff"= sips_neg - lag(sips_neg),
         "gen_diff"= sips_gen -lag(sips_gen),
         "dis_diff"= sips_dis - lag(sips_dis)) %>%
  ungroup()

summary(sips_change)

#plot longitudinal scores
ggplot(sips_change, aes(x=diff_in_days, y=sips_pos, color=bblid)) +
  geom_line() +
  labs(title="SIPS Pos")

ggplot(sips_change, aes(x=diff_in_days, y=sips_neg, color=bblid)) +
  geom_line() +
  labs(title="SIPS Neg")

ggplot(sips_change, aes(x=diff_in_days, y=sips_gen, color=bblid)) +
  geom_line() +
  labs(title="SIPS Gen")

ggplot(sips_change, aes(x=diff_in_days, y=sips_dis, color=bblid)) +
  geom_line() +
  labs(title="SIPS Dis")

#plot change
ggplot(sips_change, aes(x=pos_diff)) + 
  geom_histogram(color="white", fill= "darkgreen") + 
  labs(title="Change in SIPS Pos")

ggplot(sips_change, aes(x=neg_diff)) + 
  geom_histogram(color="white", fill= "darkgreen") + 
  labs(title="Change in SIPS Neg")

ggplot(sips_change, aes(x=gen_diff)) + 
  geom_histogram(color="white", fill= "darkgreen") + 
  labs(title="Change in SIPS Gen")
```

Looking at just change immediately pre-post 7T

``` {r sips.change.prox}
#define baseline - scan closest to 7T
#find smallest datediff that's >=365
post <- sips_scan %>%
  select(1, 2, 13:17) %>% #filter so we're not duplicating rows that wont change (demographics, 7t date, etc.)
  group_by(bblid) %>%
  filter(diff_in_days >= 365) %>%
  arrange(bblid, diff_in_days) %>%
  slice(c(1))%>% 
  rename_with( ~ paste0(.x, "_post"))

#find datediff that's <365 and closest to 0
pre <- sips_scan %>%
  group_by(bblid) %>%
  filter(diff_in_days < 365) %>%
  mutate(abs_diff_in_days=abs(diff_in_days)) %>%
  arrange(bblid, abs_diff_in_days) %>%
  slice(c(1)) %>% 
  select(!c(abs_diff_in_days)) %>% 
  rename_at(c(2, 13:17), ~ paste0(.x, "_pre"))

#new dataframe
proximal_sips <- inner_join(pre, post, by = c("bblid" = "bblid_post")) %>%
  mutate(change_pos = (sips_pos_post - sips_pos_pre),
         change_neg = (sips_neg_post - sips_neg_pre),
         change_gen = (sips_gen_post - sips_gen_pre))

#plot change
ggplot(proximal_sips, aes(x=change_pos)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Change in SIPS Pos")

ggplot(proximal_sips, aes(x=change_neg)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Change in SIPS Neg")

ggplot(proximal_sips, aes(x=change_gen)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Change in SIPS Gen")

```

## PRIME

Prime Screen scores were computed by totaling the number of items endorsed at the level of 5 or 6 and via raw sum of scores(SIP003 to SIP014) (per [Zak's paper](https://doi.org/10.1176/appi.ps.201800536) *and* [scale authors](10.1016/j.schres.2012.07.022)). Combining data from Kosha and David, keeping proband over collateral reports when one subject has both on the same day.

``` {r prime}
#add prime scores from david's spreadsheets
syrp_prime <- syrp %>%
  select(5, 7, 10, 13, 15, 17, 19, 21, 23, 25, 27, 29, 31, 33, 35) %>%
  rename(sip003=prime_1, sip004=prime_2, sip005=prime_3, sip006=prime_4, sip007=prime_5, sip008=prime_6, sip009=prime_7, sip010=prime_8, sip011=prime_9, sip012=prime_10, sip013=prime_11, sip014=prime_12, source=admin_proband) %>% #rename to match redcap prime items
  mutate(bblid=as.character(bblid)) #setting back to character for now for rbind
prime_s <- prime %>% 
  mutate(admin_proband = str_detect(redcapid, "P"), 
         admin_collat = str_detect(redcapid, "C"),
         source = case_when(admin_proband == TRUE ~ "p",
                           admin_collat == TRUE ~ "c")) %>%
    select(1, 5, 75, 8:19) %>% #subset columns
    mutate(bblid=as.character(bblid)) #setting back to character for now for rbind

#combine both prime data sources
all_prime <- rbind(syrp_prime,prime_s) %>%
   drop_na(4:15) %>% #drop rows with missing values
   mutate (bblid=as.factor(bblid)) %>% #set bblid back to factor
   distinct() #keep only unique rows

#check for multiple assessments per subj per day
id_date <- all_prime %>% select(bblid, date)
table(duplicated(id_date))

#select proband assessment if there are two proband & collateral assesments on one day
all_prime$source <- factor(all_prime$source, ordered = TRUE, levels=c("p", "c"))
prime_p <- all_prime %>%
  group_by(bblid, date) %>% #select only the first instance of a specific bblid on a specific day ~ proband
  arrange(source) %>%
  slice(c(1)) %>%
  ungroup()

#join with 7T dates & baseline dx
prime_date <- left_join(prime_p, dx_baseline, by = c("bblid" = "bblid")) 

#calc diff (negative values -> prime before 7t)
prime_date$diff_in_days = as.numeric(difftime(prime_date$date, prime_date$first_7t_date, units = "days"))

# drop data more than 1 yr prior to scan
prime_dx <- subset(prime_date, diff_in_days >= -365)
str(prime_date$bblid) #no subj lost
min(prime_date$diff_in_days) #confirm datediff filtered

#find subjects that have a baseline sips and sops up to 364 days post-7T 
clin364_pr <-  prime_dx %>% subset(diff_in_days <= 364, select = bblid) %>% unique()
#find subjects that have a followup sips sops >365 days post-7T 
clin365_pr <- prime_dx %>% subset(diff_in_days >= 365, select = bblid) %>% unique()
#filter df for appropriate timepoints
prime_scan <- subset(prime_dx, (bblid %in% clin365_pr$bblid & bblid %in% clin364_pr$bblid)) %>% 
  mutate(prime_sum = rowSums(across(4:15))) #calculate raw sum across items
  
#calculate prime_scores
prime_scan$count.5 <- apply(prime_scan[4:15], 1, function(x) length(which(x==5)))
prime_scan$count.6 <- apply(prime_scan[4:15], 1, function(x) length(which(x==6)))
prime_scan$prime_score <- rowSums(cbind(prime_scan$count.5, prime_scan$count.6))

#plot for fun
ggplot(prime_scan, aes(x=diff_in_days, y=bblid, color=prime_score)) +
  geom_point() +
  geom_vline(xintercept=365) +
  labs(title="PRIME Score Relative to First 7T Scan", x="Days from 7T Scan")

length(unique(prime_scan$bblid))
summary(prime_scan)
```

There are `r length(unique(prime_scan$bblid))` (26) subjects with PRIME scores at baseline and at least 1 post-1 yr followup. 

Look at just change immediately pre-post 7T. **TO DO: Check calculations for percent change make sense!**

``` {r prime.change.prox}
#define baseline - scan closest to 7T
#find smallest datediff that's >=365
post_prime <- prime_scan %>%
  select(1:15, 26:30) %>% #filter so we're not duplicating rows that wont change (demographics, 7t date, etc.)
  group_by(bblid) %>%
  filter(diff_in_days >= 365) %>%
  arrange(bblid, diff_in_days) %>%
  slice(c(1))%>% 
  rename_with( ~ paste0(.x, "_post"))

#find datediff that's <365 and closest to 0
pre_prime <- prime_scan %>%
  group_by(bblid) %>%
  filter(diff_in_days < 365) %>%
  mutate(abs_diff_in_days=abs(diff_in_days)) %>%
  arrange(bblid, abs_diff_in_days) %>%
  slice(c(1)) %>% 
  select(!c(abs_diff_in_days)) %>% 
  rename_at(c(2:15, 26:30), ~ paste0(.x, "_pre"))


#new dataframe
proximal_prime <- inner_join(pre_prime, post_prime, by = c("bblid" = "bblid_post")) %>%
  mutate(change_prime_sum = (prime_sum_post - prime_sum_pre),#calculate change
         change_prime = (prime_score_post - prime_score_pre),
         p.change_prime_sum = ((change_prime_sum+1e-10)/(prime_sum_pre+1e-10))*100, #calculating percent change, adding .00000000001 to each to keep from dividing by zero
         p.change_prime = ((change_prime+1e-10)/(prime_score_pre+1e-10))*100,
         p.change_prime_sum = round(p.change_prime_sum, digits=1), #rounding to undo small addition
         p.change_prime = round(p.change_prime, digits=1)) 
#calculating days between prime timepoints
proximal_prime$prime_date_diff = as.numeric(difftime(proximal_prime$date_post, proximal_prime$date_pre, units = "days"))

proximal_prime %>% select(dx, prime_sum_pre, prime_score_pre, prime_sum_post, prime_score_post, change_prime_sum, change_prime, p.change_prime_sum, p.change_prime, prime_date_diff) %>% summary()

#plots
ggplot(proximal_prime, aes(x=change_prime)) + 
  geom_histogram(color="white", fill= "darkgreen") + 
  labs(title="Change in PRIME Score")

ggplot(proximal_prime, aes(x=change_prime_sum)) + 
  geom_histogram(color="white", fill= "darkgreen") + 
  labs(title="Change in PRIME Sums")
```

Very little change in PRIME score (counts of 5 and 6 rated sxs), more variability across score sums. Negative change -> ppt's score is getting lower (i.e. symptoms improving).


Plotting PRIME with dx ~ not all points visible due to overlap

``` {r prime.change.plot}
#plot absolute change between proximal PRIME scores
ggplot(proximal_prime, aes(x=change_prime_sum, y=prime_sum_pre, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid), max.overlaps = 30) +
  labs(title="Change in PRIME Sums Over Time", x="Change (Post - Pre)", y="Baseline PRIME Total")

#plot absolute change vs time between proximal GAF Cs
ggplot(proximal_prime, aes(x=prime_date_diff, y=change_prime_sum, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid), max.overlaps = 30) +
  labs(title="Change in PRIME By Delta Time", x="Time Between PRIME Assessments (Days)", y="Change PRIME (Post - Pre)")
```

## GAF

### GAFC

Looking through SIPS GAF current 

``` {r gafc, echo=FALSE}
gafc_scan <- sips.sops_scan.filt %>% drop_na(GAF_C) %>%
  select(bblid, DOSIPS, GAF_C, first_7t_date, diff_in_days, dx, dx_source, dx_date, dx_diff)

#since dropping NA we need to refilter pre-post subj
##find subjects that have a baseline sops up to 364 days post-7T 
clin364_gafc <-  gafc_scan  %>% subset(diff_in_days <= 364, select = bblid) %>% unique()
##find subjects that have a followup sips sops >365 days post-7T 
clin365_gafc <- gafc_scan %>% subset(diff_in_days >= 365, select = bblid) %>% unique()
##filter df
gafc_scan <- subset(gafc_scan, (bblid %in% clin365_gafc$bblid & bblid %in% clin364_gafc$bblid))

length(unique(gafc_scan$bblid))
sum(is.na(gafc_scan))
```

**discordant subj counts ~ figure this out!!!*

There are `r length(unique(sops_scan$bblid))` (22) subjects with GAFC at baseline and at least 1 post-1 yr followup. 
Looking at scores across time:

``` {r gafc.plot}
#plot longitudinal scores
ggplot(gafc_scan, aes(x=diff_in_days, y=GAF_C, color=bblid)) +
  geom_line() +
  labs(title="GAFC")
```

Look at absolute change from one timepoint to the next:

```{r gafc.change}
#calculate change in GAFC scores
gafc_change <- gafc_scan %>% 
  group_by(bblid) %>% 
  arrange(diff_in_days) %>%
  mutate("gafc_diff"= GAF_C - lag(GAF_C))

summary(gafc_change)
```

Negative change -> ppt's score is getting lower (i.e. symptoms worstening).
Look at absolute change immediately pre-post 7T

``` {r gafc.change_prox}
#define baseline - scan closest to 7T
#find smallest datediff that's >=365
post_gafc <- gafc_scan %>%
  select(1:3, 5) %>% #filter so we're not duplicating rows that wont change (demographics, 7t date, etc.)
  group_by(bblid) %>%
  filter(diff_in_days >= 365) %>%
  arrange(bblid, diff_in_days) %>%
  slice(c(1))%>% 
  rename_with( ~ paste0(.x, "_post")) %>%
  ungroup()

#find datediff that's <365 and closest to 0
pre_gafc <- gafc_scan %>%
  group_by(bblid) %>%
  filter(diff_in_days < 365) %>%
  mutate(abs_diff_in_days=abs(diff_in_days)) %>%
  arrange(bblid, abs_diff_in_days) %>%
  slice(c(1)) %>% 
  select(!c(abs_diff_in_days)) %>% 
  rename_at(c(2,3,5), ~ paste0(.x, "_pre")) %>%
  ungroup()

#new dataframe
proximal_gafc <- inner_join(pre_gafc, post_gafc, by = c("bblid" = "bblid_post")) %>%
  mutate(change_gafc = (GAF_C_post - GAF_C_pre), #absolute change
         p.change_gafc = (change_gafc/GAF_C_pre)*100) #percent change
#calculating days between gafc timepoints
proximal_gafc$gafc_date_diff = as.numeric(difftime(proximal_gafc$DOSIPS_post, proximal_gafc$DOSIPS_pre, units = "days"))

#plot change
ggplot(proximal_gafc, aes(x=change_gafc)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Change in GAF Current Immediately Pre-Post 7T")

proximal_gafc %>% select(6, 3, 11, 13:15) %>% summary()
```

Plotting GAF C with dx

``` {r gafc.change.plot}
#plot absolute change vs baseline GAF C
ggplot(proximal_gafc, aes(x=change_gafc, y=GAF_C_pre, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid), max.overlaps = 30) +
  labs(title="Change in Current GAF Over Time", x="Change (Post - Pre)", y="Baseline GAF C")

#plot absolute change vs time between proximal GAF Cs
ggplot(proximal_gafc, aes(x=gafc_date_diff, y=change_gafc, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid), max.overlaps = 30) +
  labs(title="Change in Current GAF By Delta Time", x="Time Between GAF Assessments (Days)", y="Change GAF (Post - Pre)")
```

*Note: min value between GAF assessments is less than 1 yr (`r min(proximal_gafc$gafc_date_diff)` days). Will likely need to filter.*

Looking at percent change in current GAF

``` {r gafc.percent.plot}
#plot absolute change vs baseline GAF C
ggplot(proximal_gafc, aes(x=p.change_gafc, y=GAF_C_pre, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid), max.overlaps = 30) +
  labs(title="Percent Change in Current GAF Over Time", x="Percent Change (Post - Pre)", y="Baseline GAF C")

#plot absolute change vs time between proximal GAF Cs
ggplot(proximal_gafc, aes(x=gafc_date_diff, y=p.change_gafc, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = bblid), max.overlaps = 30) +
  labs(title="Percent Change in Current GAF By Delta Time", x="Time Between GAF Assessments (Days)", y="Percent Change GAF (Post - Pre)")
```

Analysis ideas  - look at improvers vs worseners? how does that this break down by diagnosis

### GASR

Looking at GASR data included in diagnosis_long.csv

``` {r gasr}
#drop missing gasr
gasr <- redcap %>% drop_na(GASR_CURRENT)%>%
  mutate(BBLID = as.factor(BBLID)) %>%
  select(1,2,6)

#join with 7T dates
gasr_date <- inner_join(gasr, dx_baseline, by = c("BBLID" = "bblid"))

#calc diff (negative values -> gasr before 7t)
gasr_date$diff_in_days = as.numeric(difftime(gasr_date$DODIAGNOSIS, gasr_date$first_7t_date, units = "days"))

# drop data more than 1 yr prior to scan
gasr_dx <- subset(gasr_date, diff_in_days >= -365)
min(gasr_dx$diff_in_days) #confirm datediff filtered

#find subjects that have a baseline sips and sops up to 364 days post-7T 
clin364_gasr <-  gasr_dx %>% subset(diff_in_days <= 364, select = BBLID) %>% unique()
#find subjects that have a followup sips sops >365 days post-7T 
clin365_gasr <- gasr_dx %>% subset(diff_in_days >= 365, select = BBLID) %>% unique()
#filter df for appropriate timepoints
gasr_scan <- subset(gasr_dx, (BBLID %in% clin365_gasr$BBLID & BBLID %in% clin364_gasr$BBLID))

#plot for fun
ggplot(gasr_scan, aes(x=diff_in_days, y=BBLID, color=GASR_CURRENT)) +
  geom_point() +
  geom_vline(xintercept=365) +
  labs(title="GASR Current Relative to First 7T Scan", x="Days from 7T Scan")

length(unique(gasr_scan$BBLID))
summary(gasr_scan)
```

There are `r length(unique(gasr_scan$BBLID))` (28) unique subjects with GASR current scores w/in 1 yr of 7T and after 1 yr follow-up.
Looking at scores across time:

``` {r gasr.plot}
#plot longitudinal scores
ggplot(gasr_scan, aes(x=diff_in_days, y=GASR_CURRENT, color=BBLID)) +
  geom_line() +
  labs(title="GASR")
```

Look at absolute change from one timepoint to the next:

```{r gasr.change}
#calculate change in GAFC scores
gasr_change <- gasr_scan %>% 
  group_by(BBLID) %>% 
  arrange(diff_in_days) %>%
  mutate("gasr_diff"= GASR_CURRENT - lag(GASR_CURRENT))

summary(gasr_change)
```

Negative change -> ppt's score is getting lower (i.e. symptoms worstening).
Look at absolute change immediately pre-post 7T

``` {r gasr.change_prox}
#define baseline - scan closest to 7T
#find smallest datediff that's >=365
post_gasr <- gasr_scan %>%
  select(1:3, 14) %>% #filter so we're not duplicating rows that wont change (demographics, 7t date, etc.)
  group_by(BBLID) %>%
  filter(diff_in_days >= 365) %>%
  arrange(BBLID, diff_in_days) %>%
  slice(c(1))%>% 
  rename_with( ~ paste0(.x, "_post")) %>%
  ungroup()

#find datediff that's <365 and closest to 0
pre_gasr <- gasr_scan %>%
  group_by(BBLID) %>%
  filter(diff_in_days < 365) %>%
  mutate(abs_diff_in_days=abs(diff_in_days)) %>%
  arrange(BBLID, abs_diff_in_days) %>%
  slice(c(1)) %>% 
  select(!c(abs_diff_in_days)) %>% 
  rename_at(c(2,3,14), ~ paste0(.x, "_pre")) %>%
  ungroup()

#new dataframe
proximal_gasr <- inner_join(pre_gasr, post_gasr, by = c("BBLID" = "BBLID_post")) %>%
  mutate(change_gasr = (GASR_CURRENT_post - GASR_CURRENT_pre), #absolute change
         p.change_gasr = (change_gasr/GASR_CURRENT_pre)*100) #percent change
#calculating days between gafc timepoints
proximal_gasr$gasr_date_diff = as.numeric(difftime(proximal_gasr$DODIAGNOSIS_post, proximal_gasr$DODIAGNOSIS_pre, units = "days"))

#plot change
ggplot(proximal_gasr, aes(x=change_gasr)) + 
  geom_bar(color="white", fill= "darkgreen") + 
  labs(title="Change in GASR Current Immediately Pre-Post 7T")

summary(proximal_gasr)
```

Plotting GASR with dx

``` {r gasr.change.plot}
#plot absolute change vs baseline GASR
ggplot(proximal_gasr, aes(x=change_gasr, y=GASR_CURRENT_pre, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = BBLID), max.overlaps = 30) +
  labs(title="Change in GASR Over Time", x="Change (Post - Pre)", y="Baseline GASR")

#plot absolute change vs time between proximal GASR
ggplot(proximal_gasr, aes(x=gasr_date_diff, y=change_gasr, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = BBLID), max.overlaps = 30) +
  labs(title="Change in GASR By Delta Time", x="Time Between GASR Assessments (Days)", y="Change GASR (Post - Pre)")
```

*Note: min value between GASR assessments is less than 1 yr (`r min(proximal_gafc$gafc_date_diff)` days). Will likely need to filter.*

Looking at percent change in GASR

``` {r gasr.percent.plot}
#plot absolute change vs baseline GASR
ggplot(proximal_gasr, aes(x=p.change_gasr, y=GASR_CURRENT_pre, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = BBLID), max.overlaps = 30) +
  labs(title="Percent Change in GASR Over Time", x="Percent Change (Post - Pre)", y="Baseline GASR")

#plot absolute change vs time between proximal GASR
ggplot(proximal_gasr, aes(x=gasr_date_diff, y=p.change_gasr, color=dx)) +
  geom_point() +
  geom_text_repel(aes(label = BBLID), max.overlaps = 30) +
  labs(title="Percent Change in GASR By Delta Time", x="Time Between GASR Assessments (Days)", y="Percent Change GASR (Post - Pre)")
```


## One Big Clinical Dataset

Make one big clinical dataset that can be subsectioned to look at different constructs but will also easily show timepoints of subjects (eg on this date, ppt X had diagnosis, SIPS and PRIME).
Merge additional clinical info (eg added by David) into our sample to potentially use as a predictor

Merging using `dx_scan_date` as our base, since it contains the 7t list & all diagnoses with 22q subj removed. 
* leave SYRP for 2nd to last since it has overlapping assessments with other dfs from Kosha (will need to re-rename prime cols)
* merge demo & dx_scan_date last and not by date so info gets applied to ALL timepoints
* calc diff_assess from 7t to date_assess

``` {r big.merge}
#list all subj so we can make sure all data is 7t subj with no 22q ~ then we can just use full joins
all_subj <- unique(dx_scan_date$bblid)

#clean cgas
summary(cgas)
cgas_clean <- cgas %>%
  mutate(bblid=as.factor(bblid),
         date_assess=as.Date(date, format="%Y-%m-%d")) %>% #clean values
  select(2, 5:8,20) %>% # select cols
  subset(bblid %in% all_subj) #make sure correct subj list
#now sips clean & merge
sips_clean <- sips_calc %>% select(1, 7:27, 58:61) %>% #sips_calc already filtered for correct source (proband>collateral) for each timepoint, frequency calculated
  rename(date_assess = DOSIPS) %>%
  subset(bblid %in% all_subj) #make sure correct subj list
#merge
merge_1 <- full_join(sips_clean, cgas_clean, by = c("bblid", "date_assess"))

#prime clean and merge (will add rest of syrp later)
names(prime_p) #version selected for correct source (proband>collateral) for each timepoint, merged prime.csv and syrp primes
prime_clean <- prime_p %>% rename(date_assess = date, prime_source=source)%>%
  subset(bblid %in% all_subj) #make sure correct subj list
#calculate prime_scores
prime_clean$count.5 <- apply(prime_clean[4:15], 1, function(x) length(which(x==5)))
prime_clean$count.6 <- apply(prime_clean[4:15], 1, function(x) length(which(x==6)))
prime_clean$prime_score <- rowSums(cbind(prime_clean$count.5, prime_clean$count.6))
prime_clean <- prime_clean %>% select(-count.5, -count.6) #remove count.5 and count.6
#merge
merge_2 <- full_join(merge_1, prime_clean, by = c("bblid", "date_assess"))

#clean dx and merge
all_dx_clean <- all_dx %>% rename(date_assess=dx_date) %>%
  mutate(bblid = as.factor(bblid)) %>%
  subset(bblid %in% all_subj) #make sure correct subj list
#merge
merge_3 <-full_join(merge_2, all_dx_clean, by = c("bblid", "date_assess"))

#clean syrp and merge ~ several proband assessments on same day, how to filter by admin_event?
syrp_clean <- syrp %>% rename(date_assess=date) %>%
  select(!c(13:37)) %>% #drop PRIME ~ maybe clean this up more later as well, lots of columns!
  subset(bblid %in% all_subj) #make sure correct subj list
#merge
merge_4 <- full_join(merge_3, syrp_clean, by = c("bblid", "date_assess"))
length(unique(merge_4$bblid)) 

#merge demographics and first scan date since they'll get applied to all clinical timepoints
first7t <- first7t %>% mutate(BBLID=as.factor(BBLID))
merge_demo.scan <- full_join(demo, first7t, by = c("BBLID")) %>%
  subset(BBLID %in% all_subj) #make sure correct subj list

#merge demo & first scan into clinical data
all_data <- full_join(merge_demo.scan, merge_4, by = c("BBLID" = "bblid"))
length(unique(all_data$BBLID)) #should always have 132 subj

#also making a version without sryp because it has multiple proband assessments per day/complicates
all_data_nosyrp <- full_join(merge_demo.scan, merge_3, by = c("BBLID" = "bblid"))
length(unique(all_data_nosyrp$BBLID)) #should always have 132 subj
#calculate datediff and filter for >1 yr pre-7t
all_data_nosyrp$date_diff = as.numeric(difftime(all_data_nosyrp$date_assess, all_data_nosyrp$first_7t_date, units = "days"))
all_data_nosyrp <- all_data_nosyrp %>% subset(date_diff >= -365)
length(unique(all_data_nosyrp$BBLID))
```

There are `r length(unique(all_data_nosyrp$BBLID))` (122) subjects who have any kind of potentially usable clinical data (e.g. not from >1 yr prior to scan).


## Appendix: Missing Diagnoses

There are `r sum(is.na(dx_scan_date$dx))` (24) subj who have 7T scans but are missing diagnoses from redcap/PNC. Writing to csv alongside info on the scan protocol and clinical data David added to Box (syrp spreadsheets) to try to figure out where these individuals are from/where more clinical data could be found.

``` {r missing.info}
#list missing dx ppts
dx_missing_info <- dx_baseline %>% subset(is.na(dx)) %>%
  select(!c(3,4,6,8:10))

#merge in syrp data to see if david has other clinical info
extra_redcap <- syrp[1:11]
dx_missing_info <- left_join(dx_missing_info, extra_redcap, by="bblid") %>%
  mutate(protocol_num = as.factor(protocol_numbe),
         redcap_id = as.factor(redcap_id),
         protocol = as.factor(protocol),
         libi_id = as.factor(libi_id),
         location = as.factor(location),
         admin_proband = as.factor(admin_proband),
         admin_event = as.factor(admin_event))
summary(dx_missing_info)
write.csv(dx_missing_info, "missing_dx.csv", row.names=FALSE, na="")
```

