---
title: "List Glucest"
output: html_document
---

Compile lists of all GluCEST scans and output as CSV to use for clinical data pull. Will compile 2 versions, one with earliest scan only (to maximize potential for clinical followup) and one including all GluCEST scans.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Install/Load packages, echo=FALSE, message=FALSE, warning=FALSE}
#load packages
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(tidyverse, ggplot2, ggpubr, ggrepel, skimr, lubridate, tidyr)
```

## Load & Clean Lists

ONM protocol - old scanner, basically same CEST sequence

To DO:
- confirm that newer scan dates = Terra 
- pull scan IDs

```{r 7t list}
glu_old <- read.csv("data/7t_dates.csv", na.strings = "") # read in 7T list from David (up to ~ summer 2019)
summary(glu_old)

# clean up formatting
glu_old <- glu_old %>%
  mutate(BBLID=as.character(BBLID),
  X7T_date = gsub(",.*","", X7T_date), 
  X7T_date = as.Date(X7T_date,format ="%m/%d/%y"),
  ONM_7T_date = as.Date(ONM_7T_date,format ="%m/%d/%y")) %>%
  rename(Terra=X7T_date,
         ONM=ONM_7T_date)
# make long
glu_old.long <- gather(glu_old, scanner, "7T_date", Terra:ONM, factor_key=TRUE) %>%
  drop_na("7T_date")
```

``` {r longglucest}
glu_ar <- read.csv("data/longglucest_scans.csv", header=T) # read in 7T list from Arianna LongGluCEST study (pulled April 2022)
summary(glu_ar)

#clean up formatting
glu_ar <- glu_ar %>%
  mutate(BBLID=as.character(BBLID),
         baseline_clinical = as.Date(baseline_clinical, format ="%m/%d/%y"),
         base_7t = as.Date(base_7t,format ="%m/%d/%y"),
         follow_7t = as.Date(follow_7t,format ="%m/%d/%y"))

# make long
glu_ar.long <- gather(glu_ar, scanner, "7T_date", base_7t:follow_7t, factor_key=TRUE) %>%
  drop_na("7T_date") %>%
  mutate(scanner= fct_collapse(scanner, Terra = c("base_7t", "follow_7t")))
```

``` {r aging}
# read in 7T list from Heather Aging study (pulled April 2022)
```

## Full Join

```{r}
#join long versions of old and Longitudinal scan lists
long1 <- merge(x=glu_old.long, y=glu_ar.long,by=c("BBLID", "7T_date", "scanner"),all=TRUE)
str(long1)

all_7t.long <- long1
```

## Wide

``` {r}
all_7t.count <- all_7t.long %>%
  group_by(BBLID) %>%
  mutate(count = as.character(row_number(BBLID))) %>%
  ungroup() # add counts
all_7t.wide <- pivot_wider(all_7t.count, id_cols=BBLID, names_from = count, values_from="7T_date") # pivot
colnames(all_7t.wide)[2:ncol(all_7t.wide)] <- paste("scan", colnames(all_7t.wide[,c(2:ncol(all_7t.wide))]), sep = "_") #rename cols

#write final list to csv
write.csv(all_7t.wide, "data/all_7T_april22.csv", row.names=FALSE, na="")
```

## First 7T

``` {r}
first_7T <- all_7t.long %>% 
  group_by(BBLID) %>%
  arrange(all_7t.long$"7T_date") %>%
  slice(1) %>%
  ungroup()
summary(first_7T)
nrow(first_7T)

#write final list to csv
write.csv(first_7T, "data/first_7T_april22.csv", row.names=FALSE, na="")
```



